---
title: "contrastanalysis"
author: "Mirka Henninger & Simone Malejka"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{contrastanalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  ```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```


### Introduction 
**contrastanalysis** allows to perform contrast analyses for independent and dependent samples. In addition to standard contrast analyses, methods are provided to test two competing hypotheses can be tested against each other. All functions and explanations are based on Rosenthal, Rosnow, and Rubin(2000) as well as Sedlmeier and Renkewitz (2008).

Contrast analyses are a sensible alternative to standard variance analyses as it allows to test specific, directed, a priori hypotheses. This leads to results that are easier to interpret and effects can be detected with a higher power.


### The first steps
After the contrastanalysis package has been installed, we load it by the following command. 

```{r}
library(contrastanalysis)
```


To illustrate the functions of this package, we use two datasets that are already available in R: a subset of variables of the **iris** data for contrast analysis for independent samples and \texttt{president} for contrast analyses for dependent samples. We load the data using the following code:

```{r, fig.height=3.5, fig.width=6}
# load data
data("iris")
iris <- iris[,c("Species", "Petal.Length")]
presidents <- data.frame(Qtr1=presidents[seq(1, length(presidents), 4)],
                         Qtr2=presidents[seq(2, length(presidents), 4)],
                         Qtr3=presidents[seq(3, length(presidents), 4)],
                         Qtr4=presidents[seq(4, length(presidents), 4)])
presidents <- na.omit(presidents)
# some overview of the data
par(mfrow = c(1,2))
plot(iris$Species, iris$Petal.Length, type = "b", main = "Example used \nfor independent samples", xlab = "Species", ylab = "Petal Length")
plot(colMeans(presidents), type = "b", main = "Example used \nfor dependent samples", xlab = "Calendar Quarter", ylab = "Approval Ratings")
```

The **iris** dataset contains the independent variable **Species** with three groups (setosa, versicolor, and virginica) and the dependent variable **Petal.Length**. The **presidents** datasets contains quarterly approval rating for the US presidents.

We show how to test a priori specified hyptheses using planned contrast analysis, but also how to directly test two competing contrasts against each other. We will give an example of each approach for independent samples using the **iris** dataset and for dependent samples using the **presidents** dataset. 


### Contrast analysis for independent samples

We would like to compare petal length of differet species of iris. We have two hypotheses:

$H_1$: An increase in petal length from setosa over versicolor to virginica

$H_2$: Setosa has smaller petal length than versicolor and virginica

For $p=3$ groups of species, we can perform $p-1=2$ planned contrasts. The contrast weights of the two contrasts are based on the two hypotheses:

$\lambda_{H_1} = (-1,0,1)$

$\lambda_{H_2} = (-2,1,1)$

Using these specification we can perform a contrast analysis for independent samples using the package. 

#### Testing planned contrasts

The contrast weights $\boldsymbol{\lambda}$ are assigned to an object using the matrix function in R with contrasts in rows and groups in columns. The data must contain subjects in rows and two columns: the first columns contains the group indicator (here **Species**), the second column contains the dependent variable (here **Petal.Length**). 

```{r}
# extract the number of groups
(nGroup <- length(levels(iris$Species)))

# define lambda weights
(lambda <- matrix(c(
  -1,0,1, # H1: An increase from setosa over versicolor to virginica
  -2,1,1), # H2: Setosa has smaller petal length than versicolor and virginica
  ncol = nGroup,
  byrow=TRUE))

# perform contrast analysis
contrast_independent(nGroup, lambda, iris)
```

The output of the **contrast_independent** function contains results for both contrasts: Sums of squares and corresponding F-values, contrast estimates and corresponding $t$-values as well as two-tailed $p$-values. Furthermore, it contains several effect sizes, such as r_{\text{effect size}} which can be interpreted as a correlation between the dependent variable and the contrast weights, $r_{\text{alerting}}$ corresponding to the correlation between group means and contrast weights, $r_{\text{alerting}}^2$ which can be interpreted similar to a determination coefficient as a measured of explained variance by the contrast $\Big(\frac{SS_{\text{contrast}}}{SS_{\text{between}}}\Big)$. $r_{\text{contrast}}$ is defined as $r_{\text{contrast}} = \sqrt{\frac{t^2}{t^2 + df}}$ and is useful for power analyses. 

The results of this empirical example show that for $\alpha = .05$, Contrast 1 and Contrast 2 fit the data well, as all $t > 46.21$ and all $p < .001$. Hence, it seems that there is an increase in petal length from setosa over versicolor to virginica, and at the same time the results suggest that setosa has smaller petal length than versicolor and virginica.


#### Testing competing contrasts against each other

However, one may not be interested in whether there is support for $H_1$ and/or $H_2$ in the data, but in *which* of the two hypotheses $H_1$ and $H_2$ outperforms the other. We can directly test the two competing hypotheses against each other: 

```{r}
# define lambda weights
lambda1 <- c(-1, 0, 1)  # H1: Iris versicolor is distinct to Iris setosa and Iris virginica
lambda2 <- c(-2, 1, 1)  # H2: Iris versicolor is more similar to Iris virginica

# Perform contrast analysis
compare_independent(nGroup=3, lambda1, lambda2, iris)
```

The output of the **compare_independent** function contains the same information as **contrast_independent** in terms of test statistics and effect sizes. But, results are shown for one test only, namely a different comparison of the two hypotheses. In addition, the standardized contrast weights as well as their difference that is used for the contrast analyses are shown in the output. 

When the $t$-value of the summary of results is positive and significantly different from $0$, the data supports Hypothesis $H_1$, while the data support hypothesis $H_2$ when the $t$-value is negative and significantly different from $0$. Hence, the results of this direct comparison show that for $\alpha = .05$, Contrast 1 does outperform Contrast 2 as $t = 2.54$ and $p = .014$. Hence, it seems that there is an increase in petal length from setosa over versicolor to virginica rather than setosa having smaller petal length than versicolor and virginica together.
The results suggest that although Iris versicolor is a hybrid of the two other species, it is very similar to Iris virginica with regards to its petal length. 


### Contrast analysis for dependent samples

We would like to compare quarterly approval rating for the US presidents and again we have two hypotheses: 

$H_1$: A decrease in approval ratings with stagnation over warmer months

$H_2$: A linear decrease in approval ratings

Please note that for $p=4$ measurement occasions, we could perform a maximum $p-1=3$ planned contrasts. However, as we only have two hypotheses, we only specify two contrast weight vectors:

$\lambda_{H_1} = (-1,0, 0, 1)$

$\lambda_{H_2} = (3, 1, -1, -3)$

Using these specification we can perform a contrast analysis for dependent samples using the package. 

#### Testing planned contrasts

As for the contrast analysis for independent samples, the contrast weights $\boldsymbol{\lambda}$ are assigned to an object using the matrix function in R with contrasts in rows and groups in columns. The data must contain subjects in rows and $p$ columns, one for each measurement occasions that contain the dependent variable. 

```{r}
(nGroup <- ncol(presidents))

# define lambda weights
(lambda <- matrix(c(
  1, 0, 0, -1, # H1: decrease in approval ratings with stagnation over warmer months
  3, 1, -1, -3), # H2: linear decrease in  approval ratings
  ncol = nGroup,
  byrow=TRUE))

# perform contrast analysis
contrast_dependent(nGroup, lambda,presidents)
```


The output of the **contrast_dependent** function contains results for both contrasts: F-values, contrast estimates and corresponding $t$-values as well as two-tailed $p$-values. Furthermore, it contains the effect size $g$, a standardized distance measure. 

The results of this empirical example show that for $\alpha = .05$, Contrast 1 and Contrast 2 fit the data well, as all $t > 3.10$ and all $p < .003$. Hence, it seems that presidental approval ratings decrease over the four calendar quarters, but it cannot be identified whether or not this decrease stagnates over warmer months. 

#### Testing competing contrasts against each other

To directly compare the two contrasts and test whether the decrease stagnates over the summer months, we directly test the two competing hypotheses against each other: 

```{r}
# define lambda weights
lambda1 <- c(1, 0, 0, -1)   # H1: decrease in approval ratings with stagnation over warmer months
lambda2 <- c(3, 1, -1, -3)  # H2: linear decrease in  approval ratings

# perform contrast analysis
compare_dependent(nGroup=nGroup, lambda1, lambda2, presidents)
```

The output of the **compare_dependent** function mirrors the information from **contrast_dependent**, but only shows results for one test only, namely a different comparison of the two hypotheses and shows the standardized contrast weights as well as their difference that is used for the contrast analyses. 

Similar to the **compare_independent* function, the data supports Hypothesis $H_1$ when the $t$-value of the summary of results is positive and significantly different from $0$, while the data support hypothesis $H_2$ when the $t$-value is negative and significantly different from $0$. Hence, the results of this direct comparison show that for $\alpha = .05$ neither of the two contrasts outperforms the other as $t = 0.53$ and $p = .59$. Hence, there is a decrease in approval ratings, but it is not decisive whether this decrease stagnates over the summer months or not. 


### References

Rosenthal, R., Rosnow, R. L., & Rubin, D. B. (2000). Contrasts and effect sizes in behavioral research: A correlational approach. Cambridge University Press.

Sedlmeier, P., & Renkewitz, F. (2008). Forschungsmethoden und Statistik in der Psychologie. München: Pearson Studium.
